Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-11-25 manuscript.qmd
continue refining the formalism

A naive approach to model the contralateral lymphatic spread would be to simply employ two independent unilateral models as introduced in @sec-unilateral. During training, one could enforce that some parameters are shared between these two models, e.g. the parameterization of the distributions over diagnose times or the spread among the LNLs ($t_{vr}). However, this approach lacks a way to describe the correlation between ipsi- and contralateral involvement discussed in section {#sec-data-strat}. Two independent unilateral models would not describe the observation that contralateral involvement becomes more likely with more severe ipsilateral involvement.

@@

Thus, we extend the formalism in @sec-unilateral in such a way that the model's ipsi- and contralateral side evolve synchronously over time. To achieve that, we start by writing down the posterior distribution of involvement an analogy to @eq-uni-bayes-law, which is now a joint probability of an involvement $\mathbf{X}^\text{i}$ ipsilaterally *and* an involvement $\mathbf{X}^\text{c}$ contralaterally, given a diagnosis of the ipsilateral LNLs $\mathbf{Z}^\text{i}$ and of the contralateral ones $\mathbf{Z}^\text{c}$:

@@

The term representing the model's prior probability of hidden involvement does not factorize. However, we assume that there is no direct lymph drainage from an ipsilateral LNL to a contralateral LNL. No major lymph vessels cross the mid-sagittal plane. In the graphical model, this means that we assume no directed arcs between ipsilateral and contralateral LNLs, i.e. tumor spread to the contralateral side is due to the primary tumor alone. We can thus write the joint probability $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right)$ as a factorising sum:

@@

This assumption makes intuitive sense: The state of the ipsilateral and contralateral sides of the lymphatic network evolves independently over time as there are no major lymph vessels crossing the midline. However, both sides are coupled via time. Qualitatively speaking, according to the model in {#eq-bilateral-marginal}, a joint state with severe contralateral involvement and limited ipsilateral involvement would be unlikely, because severe contralateral involvement becomes likely only at later time steps when limited ipsilateral involvement becomes unlikely.

@@

In general, the matrices $\boldsymbol{\Lambda}_\text{i}$ and $\boldsymbol{\Lambda}_\text{c}$ could be parameterized using a disjoint set of parameters, i.e., the ipsi- and contralateral spread rates are entirely different. However, using three sensible assumptions, we can reduce the parameter space by sharing some parameters between the two sides:

1. We assume that both ipsilateral and contralateral spread is described through the same graph shown in {#fig-full-graph}.
2. We assume the spread *among* the LNLs to be same on both sides. It is reasonable to assume the lymphatic system is symmetric. Thus, the spread rates from one LNL to the other should be symmetric, too. Formally, this means \
   $$
   \begin{aligned}
   b_v^\text{c} &\neq b_v^\text{i} \\
   t_{rv}^\text{c} &= t_{rv}^\text{i}
   \end{aligned}
   $$ {#eq-symmetries}
   for all $v \leq V$ and $r \in \operatorname{pa}(v)$.
3. The tumor's spread to the contralateral side in case of an extension over the midline is larger than if it was clearly lateralized, but smaller than its spread to the ipsilateral side. This assumption stems from a simple thought experiment: Consider moving the tumor from a clearly lateralized position accross the mid-sagittal plane to the same position, but on the contralateral side. In the beginning we would have $b_v^\text{c} < b_v^\text{i}$, while in the end, the situation is reversed. If a tumor extends over the mid-sagittal line, its contralateral spread rate can be expected to be in between these two extremes. We encode this in a *mixing parameter* $\alpha \in [0,1]$ that captures a "degree of asymmetry": \
   $$
   b_v^{\text{c},\epsilon=\texttt{True}} = \alpha \cdot b_v^\text{i} + (1 - \alpha) \cdot b_v^{\text{c},\epsilon=\texttt{False}}
   $$ {#eq-mixing}
   This means the model now uses three different sets of parameters to describe the spread from the tumor to the LNLs: $b^\text{i}_v$ for the spread to the ipsilateral LNLs, $b_v^{\text{c},\epsilon=\texttt{False}}$ for the spread to the contralateral LNLs as long as the tumor is clearly lateralized, and finally $b_v^{\text{c},\epsilon=\texttt{True}}$ when it crosses the midline. Note, however, that these three sets of spread rates only account for $2 \cdot 2^V + 1$ parameters, since they are coupled via the mixing parameter $\alpha$.

Our parameter space has now expanded to

@@

which is less than double in size, compared to the unilateral model. From these parameters, we now build three different transition matrices: The unchanged matrix $\mathbf{A}_\text{i}$ for the ipsilateral side, one $\mathbf{A}_\text{c}^{\epsilon=\texttt{False}}$ for the contralateral side that covers the progression as long as the tumor is lateralized, and $\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}$ in case of a tumor that has crossed the mid-sagittal line.

@@

It can be assumed that most tumors that cross the midline at the time of diagnose have started as lateralized tumors and grew over the midline at a later time step. Hence, the transition matrix $\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}$ only applies for a subset of the time steps.
Therefore, we also model the tumor's extension over the mid-sagittal line as a binary random variable. A tumor starts lateralized and at every time-step there is a finite probability $p_\epsilon$ that the tumor grows over the midsaggital plane. The overall probabilities to find a patient with a clearly lateralized tumor or one that extends over the mid-sagittal line after $t$ time-steps are then respectively given by

@@

The case when midline extension is eventually present is more complicated: We already marginalize over the exact time-step when the tumor grows over the mid-sagittal line. But whenever that happens, we also need to change the contralateral transition matrix to use the increased spread rates $b_v^{\text{c}, \epsilon=\texttt{True}}$ from the tumor to the contralateral LNLs, given by the linear mixing in @eq-mixing. We can achieve the correct marginalization by iteratively building the joint distribution $P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t \right)$. We start from $t=0$, where we know that all LNLs are healthy (the contralateral neck is in the state $\mathbf{X}_\text{c}=\boldsymbol{\xi}_0$) and the tumor is lateralized ($\epsilon=\texttt{False}$):

@@

whereas all other states have zero probability. Then we consider an arbitrary later time-step $t=\tau+1$. There are two possible scenarios we need to marginalize over:

1. The tumor was still lateralized at $t=\tau$ and just grew over the midline. Under this scenario, the probability for having a midline extension at $t=\tau+1$ is given by $p_\epsilon$. We use this to weight the contralateral state distribution that was so far evolved without increased contralateral spread.
2. The mid-sagittal line was already crossed by the tumor. In that case, the probability is 1 to remain in that lateralization state. Thus, to account for this case we simply add the distribution $P\left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau \right)$ from one time-step earlier, arriving at a recursion relation:

@@

We can use this term to compute the likelihood of all patients with and without midline extension separately. And if for some patients the information of tumor lateralization is not available, we can simply sum the above term once for $\epsilon=\texttt{False}$ and once for $\epsilon=\texttt{True}$ to marginalize over the unknown variable.

>>>>

A straightforward approach to modeling contralateral lymphatic spread would be to use two independent unilateral models, as described in Section @sec-unilateral, with shared parameters such as the distribution of diagnosis times or spread between LNLs ($t_{vr}$). However, this method would fail to capture the correlation between ipsilateral and contralateral involvement discussed in @sec-data-strat, particularly the observed increase in contralateral involvement with greater severity of ipsilateral spread.

@@

Thus, we extend the formalism in @sec-unilateral in such a way that the model's ipsi- and contralateral side evolve synchronously over time. To achieve that, we start by writing down the posterior distribution of involvement in analogy to @eq-uni-bayes-law, which is now a joint probability of an involvement $\mathbf{X}^\text{i}$ ipsilaterally *and* an involvement $\mathbf{X}^\text{c}$ contralaterally, given a diagnosis of the ipsilateral LNLs $\mathbf{Z}^\text{i}$ and of the contralateral ones $\mathbf{Z}^\text{c}$:

@@

The term representing the model's prior probability of hidden involvement does not factorize. However, we assume no direct lymphatic drainage from ipsilateral to contralateral LNLs, as major lymph vessels do not cross the mid-sagittal plane. In the graphical model, this translates to the absence of directed arcs between ipsilateral and contralateral LNLs, implying that contralateral tumor spread occurs solely via the primary tumor. We can thus write the joint probability $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right)$ as a factorising sum:

@@

This assumption is intuitive: since no major lymph vessels cross the midline, the ipsilateral and contralateral sides of the lymphatic network evolve independently over time. However, they are indirectly coupled through time. For example, a joint state with severe contralateral involvement and limited ipsilateral involvement is improbable: Severe contralateral involvement typically occurs at later time steps, when limited ipsilateral involvement is unlikely.

@@

The matrices $\boldsymbol{\Lambda}\text{i}$ and $\boldsymbol{\Lambda}\text{c}$ could, in principle, be parameterized with entirely separate parameters, allowing ipsilateral and contralateral spread rates to differ completely. However, we simplify the parameter space by sharing parameters between the two sides, based on the following three assumptions:

1. **Shared Graph Structure**: Both ipsilateral and contralateral spread are described by the same graph shown in @fig-full-graph.
2. **Symmetric Spread Among LNLs**: The spread among LNLs is assumed to be the same on both sides, reflecting the symmetric structure of the lymphatic system. Consequently, the spread rates between nodes should also be symmetric. This is formalized as:\
  $$
  \begin{aligned}
  b_v^\text{c} &\neq b_v^\text{i} \\
  t_{rv}^\text{c} &= t_{rv}^\text{i}
  \end{aligned}
  $$ {#eq-symmetries}
  for all $v \leq V$ and $r \in \operatorname{pa}(v)$.
3. **Asymmetric Spread from Tumor**: Tumor spread to the contralateral side varies depending on whether the tumor crosses the mid-sagittal plane. When the tumor extends over the midline, its spread rate to the contralateral LNLs should be larger than if the tumor were lateralized, but still smaller than the tumor's spread to the ipsilateral side. To model this, we introduce a mixing parameter $\alpha \in [0,1]$, representing the "degree of asymmetry" effectively mixing these two extremes:\
  $$
  b_v^{\text{c},\epsilon=\texttt{True}} = \alpha \cdot b_v^\text{i} + (1 - \alpha) \cdot b_v^{\text{c},\epsilon=\texttt{False}}
  $$ {#eq-mixing}
  This parameterization results in three sets of rates for tumor spread to the LNLs: (1) the spread to ipsilateral LNLs $b^\text{i}_v$, (2) the spread to contralateral LNLs as long as the tumor is lateralized $b_v^{\text{c},\epsilon=\texttt{False}}$, (3) the spread to contralateral LNLs when the tumor crosses the midline $b_v^{\text{c},\epsilon=\texttt{True}}$. Despite this additional complexity, these rates collectively account for only $2 \cdot 2^V + 1$ parameters due to their coupling via $\alpha$.

The parameter space now becomes:

@@

This results in less than a doubling of parameters compared to the unilateral model. From these parameters, we construct three transition matrices: the unchanged $\mathbf{A}_\text{i}$ for the ipsilateral side, $\mathbf{A}_\text{c}^{\epsilon=\texttt{False}}$ for contralateral progression while the tumor is lateralized, and $\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}$ for cases where the tumor crosses the mid-sagittal plane.

@@

Most tumors crossing the midline at the time of diagnosis likely began as lateralized tumors that grew over the midline at a later stage. As a result, the transition matrix $\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}$ applies only to a subset of time-steps.

To account for this, we model the tumor's extension over the mid-sagittal line as a binary random variable. A tumor starts as lateralized, with a finite probability $p_\epsilon$ at each time step of crossing the midline. The overall probabilities of a patient having a clearly lateralized tumor or one extending over the mid-sagittal line after $t$ time steps are then given by

@@

The case of midline extension is more complex: we already marginalize over the exact time step when the tumor grows over the mid-sagittal line. However, at the point of crossing, the contralateral transition matrix must switch to the increased spread rates, $b_v^{\text{c}, \epsilon=\texttt{True}}$, as defined by the linear mixing in @eq-mixing. To correctly perform this marginalization, we iteratively construct the joint distribution $P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t \right)$.

We begin at $t=0$, where all contralateral LNLs are healthy (i.e., $\mathbf{X}_\text{c}=\boldsymbol{\xi}_0$) and the tumor is lateralized ($\epsilon=\texttt{False}$):

@@

while all other states have zero probability.

At some later time step $t=\tau+1$, there are two scenarios to marginalize over:

1. **The tumor was lateralized at $t=\tau$ and grew over the midline at $t=\tau+1$:**  
   In this case, the probability of midline extension at $t=\tau+1$ is $p_\epsilon$. This probability weights the contralateral state distribution that had previously evolved without increased contralateral spread.

2. **The tumor had already crossed the midline before $t=\tau$:**  
   Here, the tumor remains in the midline-crossed state with probability 1. To account for this scenario, we simply include the distribution $P\left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau \right)$ from the previous time step.

Combining these scenarios leads to a recursive formulation:

@@

With the above, we compute the likelihood of all patients with and without midline extension separately. And if for some patients the information of tumor lateralization is not available, we can simply marginalize over the unknown variable $\epsilon \in \{ \texttt{False}, \texttt{True} \}$.
